The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-08-16 paper/paper.Rnw
Downloaded all data from pre-primary

results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))

# results_math <- read_rds("./data/delete.Rdata")
# results_read <- read_rds("./data/delete_read.Rdata")
# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
# US is missing for reading

@@

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer("name") %>%
  ggplot(aes(name, value)) +
  geom_col() +
  scale_y_continuous(name = "Standard deviation of raw test scores") +
  scale_x_discrete(name = NULL) +
  ggtitle("Standard deviation of test scores over time") +
  theme_few() +
  theme(text = element_text(size = 14))
@
\caption{Standard deviation of test scores across all waves.}
\end{center}
\end{figure}

@@

summary_data <-
  map(escs_data, edu_calc) %>%
  setNames(years) %>%
  enframer()
@

@@

print(table_coming,
      scalebox = '0.75',
      floating = FALSE,
      add.to.row = addtorow,
      include.colnames = FALSE,
      include.rownames = FALSE,
      hline.after = nrow(table_coming),
      caption.placement = "top",
      caption = "SES sample size and ISCED composition")
@

@@

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.075, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES is catching up",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@
\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
\end{center}
\end{figure}

@@

data_to_plot %>%
  ggplot(aes(scale(Mean), difference)) +
  geom_point() +
  geom_smooth()
@

@@

>>>>

# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))

results_math <- read_rds("./paper/data/delete.Rdata")
results_read <- read_rds("./paper/data/delete_read.Rdata")
results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
# US is missing for reading

@@

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer("name") %>%
  ggplot(aes(name, value)) +
  geom_col() +
  scale_y_continuous(name = "Standard deviation of raw test scores") +
  scale_x_discrete(name = NULL) +
  ggtitle("Standard deviation of test scores over time") +
  theme_few() +
  theme(text = element_text(size = 14))

rm(var_years)
@
\caption{Standard deviation of test scores across all waves.}
\end{center}
\end{figure}

@@

summary_data <-
  map(escs_data, edu_calc) %>%
  setNames(years) %>%
  enframer()

rm(escs_data)
@

@@

print(table_coming,
      scalebox = '0.75',
      floating = FALSE,
      add.to.row = addtorow,
      include.colnames = FALSE,
      include.rownames = FALSE,
      hline.after = nrow(table_coming),
      caption.placement = "top",
      caption = "SES sample size and ISCED composition")

rm(sample_tables_topbottom, sample_tables_topmid, sample_tables_midbottom,
   avg_performance, summary_table, table_coming)
@

@@

<<deleting_used_data>>=
rm(diff_data, us_math_2000, us_math_2000_upper, us_math_2000_lower)
@

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.075, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES is catching up",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()

rm(full_data, avg_sd_increase_high, avg_sd_increase_low)
@
\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
\end{center}
\end{figure}

@@

data_to_plot %>%
  ggplot(aes(scale(Mean), difference)) +
  geom_point() +
  geom_smooth()

rm(country_scores)
@

@@

<<>>=

# 2000
# SC03Q01 School public/private
# 1 Public
# 2 Private
# 7 N/A
# 8 M/R
# 9 Mis

# 2003
# SC03Q01 (8) Public or private
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Miss

# 2006
# SC02Q01 (8) Public or private
#  1 Public
#  2 Private
#  7 N/A
#  8 Invalid
#  9 Missing

# 2009
# SC02Q01
# 1 public
# 2 private

# 2012
# SC01Q01 Public or private Num
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Missing

# 2015
# SC013Q01TA
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Missing

# 2000: ST25Q01
# 1 <ISCED 2A>
# 2 <ISCED 2B>
# 3 <ISCED 2C>
# 4 <ISCED 3A>
# 5 <ISCED 3B>
# 6 <ISCED 3C>
# 7 N/A
# 8 M/R
# 9 Mis

# 2003: ISCEDO
# 2006: ISCEDO
# 2009: ISCEDO
# 2012: ISCEDO
# 2015: ISCEDO

# 1	General
# 2	Pre-Vocational
# 3	Vocational
# 4	Modular

options(tibble.max_extra_cols = 10)
library(WDI)

wd_extract <- function(var, name) {
  wdi_data <-
    WDI(indicator = var, start = 2000, end = 2015) %>%
    select(-1) %>%
    as_tibble() %>%
    mutate(year = as.character(year))

  names(wdi_data)[2] <- name
  wdi_data
}

pop_preprimary <- wd_extract("SP.PRE.TOTL.IN", "pop_preprimary")
teacher_pupil_ratio <- wd_extract("SE.PRE.ENRL.TC.ZS", "teacher_pupil_ratio")
perc_teacher_trained <- wd_extract("UIS.TRTP.0", "perc_teach")
preprimary_priv_enrol <- wd_extract("SE.PRE.PRIV.ZS", "preprimary_priv_enrol")

difference_data <-
  complete_data_topbottom %>%
  filter(type_test == "math") %>%
  select(wave, country, difference) %>%
  rename(year = wave) %>%
  mutate(year = as.character(year))

library(inequalityintsvy)

difference_data %>%
  left_join(primary_enrolment) %>%
  left_join(pop_tot) %>%
  lmer(difference ~ enrolment + (1 | country), data = .) %>%
  arm::display()
